Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-09 ms.tex
inner stability of equilibrium

\usepackage{amssymb,amsmath,amsthm}
\usepackage{longtable}
\usepackage[hyphens]{url}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
%\usepackage{nameref} % included in hyperref
\usepackage[auth-sc]{authblk}

@@

Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{w}^* v = (I-C+MC)D v.
\end{equation}
The existence of the leading eigenvalue $\bar{w}^*$ and its corresponding right eigenvector $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix
(the leading eigenvalue $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}).

@@

@@

\bibliographystyle{agsm}
%\bibliography{/Users/yoavram/Documents/library}
\bibliography{ms}

>>>>

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage{longtable}
\usepackage[hyphens]{url}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
%\usepackage{nameref} % included in hyperref
\usepackage[auth-sc]{authblk}

@@

Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{w}^* v = (I-C+MC)D v.
\end{equation}
Note that (i) the existence of $\bar{w}^*$ and $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix; 
(ii) the local stability of this equilibrium is proven in~\nameref{sec:AppD}; and
(iii) $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}.

@@

% Appendix D: inner stability
\subsection*{Appendix D}\label{sec:AppD}
\begin{lemma}[Stability of the equilibrium $v$]
The equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is locally stable.
In addition,
\begin{enumerate}
\item If $M$ is irreducible and $D$ and $C$ are positive, then the equilibrium is globally stable.
\item If $M$ is irreducible and either $D$ or $C$ are non-negative but not strictly positive, then the equilibrium is globally stable except possibly for initial conditions on the boundary, i.e., when starting with zero elements in the initial frequency vector.
\end{enumerate}
\end{lemma}
\begin{proof}
Set $A=(I-C+MC)D$ and rewrite Eq.~\ref{eq:model} such that $f(t)$ is the frequency vector at time $t$ and
\begin{equation}
f(t) = \frac{Af(t-1)}{\norm{Af(t-1)}},
\end{equation}
where $\norm{f(t)} = \sum_{k=1}^{n}f_k(t)$.
Denote the eigenvalues of $A$ as $\lambda_1, \ldots, \lambda_n$ such that $\abs{\lambda_1} \ge \ldots \ge \abs{\lambda_n}$ and $\lambda_1$ is the \emph{Perron} leading eigenvalue, that is, the largest real eigenvalue with corresponding real non-negative right eigenvector $v$ ($\lambda_1$ is shown in~\nameref{sec:AppA} to be the stable population mean fitness $\bar{w}^*$).

First, note that the normalization $x \goesto \frac{1}{\norm{x}}$ can be applied once instead of $t$ times~\citep[Appendix~A]{Wakano2004}, that is,
\begin{equation}
f(t) = \frac{A^t f(0)}{\norm{A^t f(0)}}.
\label{eq:ftplus1}
\end{equation}

We write $f(0)$ as a linear combination of eigenvectors $v^k$ of $A$ for $k=1, \ldots, n$:
\begin{equation}
f(0) = a_1 v^1 + a_2 v^2 + \ldots + a_n v^n,
\label{eq:linear_combination}
\end{equation}
such that the eigenvector corresponding to the leading eigenvalue is $v=v_1$.

To proceed we need to make sure that $a_1 \ne 0$.
If $M$ is reducible then $A$ is reducible. 
In this case, $a_1 \ne 0$ can be guaranteed by setting $f(0)=v + \epsilon$ such that $\epsilon_k$ is small enough and $v_k+\epsilon_k \ge 0$ for all $k$, and $\sum_{k=1}^{n}{\epsilon_k} = 0$.

If $M$ is irreducible and $D$ and $C$ are positive, then $A$ is irreducible and  any choice of $f(0)$ guarantees $a_1 \ne 0$.

If $M$ is reducible and either $D$ or $C$ are non-negative but not strictly positive, than any choice of $f(0)$ such that $f_k(0)>0$ for all $k$ guarantees $a_1 \ne 0$.

We write, using Eq.~\ref{eq:linear_combination},
\begin{multline}
A^t f(0) = 
a_1 A^t v^1 + a_2 A^t v^2 + \ldots + a_n A^t v^n = \\
a_1 \lambda_1^t v^1 + a_2 \lambda_2^t v^2 + \ldots + a_n \lambda_n^t v^n = \\
\lambda_1^t \bigg(a_ 1v^1 + a_2 \bigg(\frac{\lambda_2}{\lambda_1}\bigg)^t v^2 + \ldots +  a_n \bigg(\frac{\lambda_n}{\lambda_1}\bigg)^t v^n\bigg).
\end{multline}
Therefore,
\begin{equation}
\frac{A^t f(0)}{\lambda_1^t} \goesto a_1 v_1
\end{equation}
as $t \goesto \infty$, because $\abs{\frac{\lambda_k}{\lambda_1}} < 1$
for all $k=2, \ldots, n$.
Finally, 
\begin{equation}
f(t) = 
\frac{A^t f(0)}{\norm{A^t f(0)}} =
\frac{\frac{A^t f(0)}{\lambda_1^t}}{\norm{\frac{A^t f(0)}{\lambda_1^t}}} \goesto
\frac{a_1 v^1}{\norm{a_1 v^1}} = v,
\end{equation}
as $t \goesto \infty$, which completes the proof.
\end{proof}

@@

\bibliographystyle{agsm}
\bibliography{/Users/yoavram/Documents/library}
%\bibliography{ms}
